Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-09-08 output/breed_paper.Rnw
Latest edits

\title{Rice yield improvements by plant breeding are offset by
  decreases over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
\author[ucd]{David MacKill}
\author[ucd]{Jim E. Hill}
\author[res]{Kent McKenzie}
\author[ucd]{Chris van Kessel}
\author[ucd]{Bruce A. Linquist}

\address[ucd]{Dept. of Plant Sciences, University of California --
Davis, Davis CA 95616}

\address[res]{Rice Experiment Station, California Cooperative Rice
Research Foundation, Biggs, CA 95917}

@@

Meeting the challenge of feeding an growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite demonstrated increased yield
potential of new genetics over old in side-by-side comparisons. We
analyzed experimental data from one such system, California (CA) rice
production, in order to better understand what improvements have been
made in the genetic yield potential via plant breeding. Specifically,
the hypothesis that a rice variety's yield erodes over time after the
initial selection was tested as an explanation for how side-by-side
comparisons show increased yield while no substantive increases have
been seen in farmers' fields. Side-by-side comparisons under yield
erosions will show ``false'' increases due to the older cultivar
decreasing, not due to increased genetic yield potential by the newer
cultivar. We confirmed an apparent increase in yield in side-by-side
comparisons between old and newer varieties when the yield erosion was
not considered. However, we demonstrate this apparent increase in
yield can be explained by the differences in the time since selection
between cultivars. The impact of yield erosion was estimated to be
29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible interal -4.4 to
-53.3) after initial selection. Once this effect was considered, the
apparent yield advantage of newer cultivars over old was uncertain
(-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval -36.1 to
31.5). This underscores the importance of breeding to maintain yield
and the challenges of increasing the genetic yield potential.

@@

\begin{keyword} rice, yield potential, breeding
\end{keyword}

@@

Constraints on arable land are increasing simultaneous with the need
to increase total production to meet growing demand \citep{Foley2011a,
Godfray2010, Mueller2012}, which has necessitated harvesting more
grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple crops,
however more and more production systems are experiencing plateaus in
grain yield \citep{Grassini2013, Peng2010}. If continued improvements
per unit land area are no longer possible, increasing the area under
cultivation will be needed to meet demand, which carries with it
undesirable ecological implications. Therefore, it is of critical
importance to better understand why grain yields have slowed or
plateaued in intensified production systems.

California (CA) rice production represents one such production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly (Fig. \ref{fig:state_yield}), in part due to
improvements in rice genetics (i.e., the ``Green Revolution'') through
the efforts of the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). However, since 1990 CA state average yields have increased
more slowly (Fig. \ref{fig:state_yield}) despite continuous plant
improvement \citep{ccrrf}. For the last 30 years, each newly developed
cultivar has demonstrated a yield advantage over the previous
cultivars in the Statewide Variety Trials (an annual multi-site
variety trial program that compares experimental entries side-by-side
with ``check'' cultivars). Yet, taken across years, these same reports
show stagnant yields. For example, at the year of release (1985),
cultivar M-202 averaged 11 Mg ha\tsups{-1} \citep[14\%
MC;][]{m202}. In 2015, the reported average yield of the newest released
cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep{m209}. Yet this same
report, M-209 demonstrated a X\% yield advantage over M-202. In other rice
systems, a decrease in a cultivar's yield performance over time has
been observed \citep{Peng1998,
Peng2010, Peng2000, Datta1995}, often termed yield erosion
\citep{Peng1998}. It has been speculated that this is due to the
inability of a selected cultivar's genetics to adapt to changing  biotic and
abiotic conditions. We sought to quantify this phenomenon to better
understand how plant improvement and yield erosion are contributing to
the slow yield improvement in farmers' fields over the last
30 years in CA. Specifically, two competing hypotheses were
tested: (1) the genetic yield advantage has increased over the last
30+ years, and (2) the genetic yield advantage has not increased over
the last 30+ and the erosion of yield over
time is creating the illusion of improved yields (i.e., the new
cultivar is recovering the old benchmark as the older cultivar's yield
erodes).

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
top 15cm. As the central location for the CCRRF breeding program, the RES is
researcher-managed (as opposed to farmer-managed trials) and conducts
the bulk of the trials for the Statewide Variety Trials (approximately
3 to 4 times the number of plots compared to any other location). For
these reasons, we focused our investigation on the RES to avoid
potential complications due to site, climate, and management
differences at the other 6 to 8 farmer-managed trials per year.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against named and released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute roughly
90\% of the planted area in CA. These cultivars are typcically planted
in the Statewide Variety Trials for 2 to 4 years prior to their
official release. For these cultivars planted in
breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations were
collected leading to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release (typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability) and the
time elapsed (in years) since release was calculated.

Plots at the RES are managed similar to University of California
Cooperative Extension prescribed best management
practice, including land preparation, fertility management, and pest
and disease control \citep{ucce}. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

Yield was modeled as a response to the number of years since the
variety was officially released and the year of release using a
mixed-effects linear regression. To account for similarities between years
and between cultivars, year and cultivar were included in the model as
random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across varieties and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was as above except omitting the $Rate_{yield\ erosion\ i}$ term.

@@

Data was processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval \citep{rstanarm}. Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

@@

@
\caption{The performance of two major cultivars at the CA Rice
Experiment Station (RES) over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M-202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs}
\end{figure}

@@

The full model which corrected for both the year of release and yield
erosion was estimated to have better out-of-sample predictive accuracy
compared to reduced models containing only one predictor. As estimated
by the best model, there is no conclusive evidence that the yields of
medium grain rice cultivars selected and released by CCRRF's breeding
program have, on average, increased over the last 30 years. The
estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:w_erosion}). Hence, the evidence here is consistent
with little to no increase in yield potential over the study
period. When yield erosion is not corrected for in the comparison
(i.e., omitted from the model), there is an apparent small increase
per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}), though this
estimate ignores the confounding effect of yield erosion over time and
performed worse in out-of-sample predictive accuracy.

After correcting for the other variables, some cultivars are estimated
to have greater yield compared to the overall average. The marginal
estimate for the performance of a cultivar averaged over the
uncertainty in the other parameters in the model (i.e., the random
effect) shows three cultivars with higher than average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:w_erosion} \& \ref{fig:ids_comp}). A fourth, M-209,
might also have higher than average yield, though the CI for M-209
overlaps with zero (Fig. \ref{fig:ids_comp}).

\begin{comment}
\begin{figure}[H]

@@

@
\caption{The relationship between yield and the release year of medium
grain rice cultivars developed by the California Cooperative Rice
Research Foundation (California, USA), for the last 6 years
(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
correcting for yield erosion over time.}
\label{fig:wo_erosion}
\end{figure}
\end{comment}

\begin{figure}[H]
<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=

@@

@
% Change from a./b. to accounting for erosion, More details about how
% yield was corrected - 
\caption{The relationship between yield and the release year of medium
grain rice cultivars developed by the California Cooperative Rice
Research Foundation (California, USA), for the last 6 years
(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
correcting for yield erosion over time. Yield erosion was corrected
for by adjusting each observed yield by estimated amount lost 
  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
\label{fig:w_erosion}
\end{figure}

@@

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is estimated to be
  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
  and bars (|) represent the mean and standard error of the raw data,
  respectively. The shaded area is the 90\% credible interval.}
% For each year, point size by n (open circles)
\label{fig:RES_year}
\end{figure}

There is estimated to be an decrease of yield over time once a
cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}, 90\% CI: \Sexpr{RES_CI[2]} to
\Sexpr{RES_CI[1]}). The oldest cultivar in the trials is estimated to
have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since release. Comparing two dominate medium grain rice
cultivars M-202 (widely planted from the late 1980s until the early
2000s), and M-206 (the most widely planted CCRRFS cultivar at
approximately 50\% of planted area at the time of writing), the two
cultivars experienced similar trends over time to each other
(Fig. \ref{fig:compare_yrs}) and to the estimate from all released
medium grains (Table \ref{tbl:coef}). At the time of writing, an
estimated \Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
over M-202 can be attributed solely to the erosion of M-202's yield
over time (Figs \ref{fig:w_erosion} \& \ref{fig:compare_yrs}). Further
supporting this estimate, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:compare_yrs}, Table \ref{tbl:ids}).

@@

We estimate that yield erosion has decreased the yield of medium
grain rice cultivars released by the CCRRF over the last 30+ years by
an average of \Sexpr{yrs_coef} Mg ha\tsups{-1},
potentially offsetting any gains made by the breeding program. Biotic
and abiotic forces have previously been speculated to be responsible
for yield erosion, since the cultivar's genetics remain stable from the
time of selection but the conditions the cultivar may experience over
time are continuously changing \citep{Peng1998, Peng2000,
Peng2010}. The trend observed here is consistent with previous reports
of the erosion of genetic gains of cultivar IR8 at IRRI
\citep{Peng1998, Peng2000}, suggesting the erosion of yield over time
is a phenomenon that widely affects rice breeding programs. This
highlights both the need for maintenance breeding and the difficulty
of increasing genetic yield potential, especially in systems such as
CA where yield is approaching the physiological limit
\citep{espe16a}. Without maintenance breeding continuously selecting
cultivars well adapted to current conditions, CA rice production
systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
the study period.

Overall, we do not find conclusive evidence of increased genetic yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons for
yield erosion over time, the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
year\tsups{-1}. This suggests that both increases and decreases in the
genetic yield potential over the study period. However, even when the
comparison is not corrected for yield erosion, the estimated
improvements over time (\Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; Fig. \ref{fig:wo_erosion}) are still much less than
previous estimates of 37 to 60 kg ha\tsups{-1} year\tsups{-1}
\citep{Omar2014}. We conclude that the breeding program has likely
been maintaining but not substantially improving yield potential in
the face of changing biotic and abiotic pressures.

>>>>

\title{Rice yield improvements through plant breeding are offset by
  inherent yield declines over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
\author[ucd]{David MacKill}
\author[ucd]{Jim E. Hill}
\author[res]{Kent McKenzie}
\author[ucce2]{Michelle Leinfelder-Miles}
\author[ucce]{Luis A. Espino}
\author[ucce]{Randall Mutters}
\author[ucd]{Chris van Kessel}
\author[ucd]{Bruce A. Linquist}

\address[ucd]{Dept. of Plant Sciences, University of California --
Davis, Davis CA 95616, USA}

\address[res]{Rice Experiment Station, California Cooperative Rice
Research Foundation, Biggs, CA 95917, USA}

\address[ucce]{University of California Cooperative Extension,
Oroville, CA 95965, USA}

\address[ucce2]{University of California Cooperative Extension,
Stockton, CA 95206, USA}

@@

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
better understand what improvements have been made in the genetic
yield potential obtained through plant breeding. Specifically, the
hypothesis was tested that the yield of a rice variety declines over
time after the initial selection was tested. This would explain how
side-by-side comparisons show increased yield while no substantive
increases have been seen in farmers' fields. Side-by-side comparisons
under yield erosion will show ``false'' yield increases due to the
yield of the older cultivar decreasing, not due to the increased
genetic yield potential of the newer cultivar. An apparent increase in
yield in side-by-side comparisons between old and new rice varieties
was observed when the yield erosion was not considered. However, this
apparent increase in yield could be explained by the differences in
the time when these cultivars were selected. The impact of yield
erosion was estimated to be 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
credible interval -4.4 to -53.3) after initial selection. Once this
effect was considered, the apparent yield advantage of newer cultivars
over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
credible interval -36.1 to 31.5). This underscores the importance of
breeding to maintain yield and the challenges of increasing the
genetic yield potential.
% Add more implications

@@

\begin{keyword} rice, yield potential, breeding, yield erosion
\end{keyword}

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However production systems are experiencing plateaus in grain
yield \citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications. Therefore, it is of critical importance to better
understand why the rate of increase in grain yields has declined or
leveled off in intensified production systems.

California (CA) rice production represents one such production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly (Fig. \ref{fig:state_yield}), in part due to
improvements in rice genetics (i.e., the ``Green Revolution'') through
the efforts of the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). However, since 1990 CA state average yields have increased
more slowly (Fig. \ref{fig:state_yield}) despite continuous crop improvement \citep{ccrrf}. For the past 30 years, each newly developed
cultivar has demonstrated a yield advantage over the previous
cultivars in the Statewide Variety Trials (an annual multi-site
variety trial program that compares experimental entries side-by-side
with ``check'' cultivars). However, in a seeming contradiction these same
reports show stagnant yields across release announcements (i.e.,
reports accompanying the office release) from the
various years. For example, at the year of release
(1985), cultivar M-202 was reported to yield an average of 11 Mg
ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the reported average
yield of the newest released cultivar M-209 was 10.8 Mg ha\tsups{-1}
\citep]14\% MC;][]{m209}. In other rice systems, a decrease in a
cultivar's yield performance over time has been observed
\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
erosion \citep{Peng1998}. It has been speculated that this yield erosion is due to
the inability of a selected cultivar's genetics to adapt to changing
biotic and abiotic conditions. To better understand how
plant improvement and yield erosion are contributing to the
improvement of rice yield in farmers' fields, we teseted two competing hypotheses: (1) the
genetic yield advantage increased over the last 30+ years, and (2)
the genetic yield advantage has not increased over the last 30+
because the erosion of yield over time creates an illusion of improved
yields.

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
breeding program, the RES is researcher-managed (as opposed to
farmer-managed trials) and conducts the majority of the trials for the
Statewide Variety Trials (approximately 3 to 4 times the number of
experimental plots compared to any other location). To avoid potential
complications due to site, climate, and management differences at the
other 6 to 8 farmer-managed trials per year we focused our
investigation on the RES.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially named and released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute approximately
90\% of the planted rice area in CA. These cultivars are typcically planted
in the Statewide Variety Trials for 2 to 4 years prior to their
official release. For these cultivars planted in
breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations were
collected leading to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations representing
\Sexpr{length(unique(mod_data[,'id']))} cultivars,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release, typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability and the
time elapsed (in years) since release was calculated.

Experimental plots at the RES are managed similar to University of California
Cooperative Extension following prescribed best management
practice, including land preparation, fertility management, and pest
and disease control \citep{ucce}. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

To test each hypothesis, the influence of breeding and yield erosion
were quantified on yield over time. Yield was modeled as a response to
the number of years since the variety was officially released and the
year of release using a mixed-effects linear regression. To account
for similarities between years and between cultivars, year and
cultivar were included in the model as random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across varieties and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was used as described above except omitting the $Rate_{yield\ erosion\ i}$ term.

@@

Data were processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval \citep{rstanarm}. Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

@@

\begin{figure}[H]
<<fig=TRUE, width = 6, height = 5, out.width = '85%', echo=FALSE>>=

idx = mod_data[,"yrs_in_trial"] %in% -2:0

plot_data = prep_plot_data(mod_data[idx,"yield_kg"] / 1000 ,
                           mod_data[idx, "id"])

plot_data[,"yr"] = tapply(mod_data[idx, "release_yr"],
                          mod_data[idx,"id"], unique)

plot(mean ~ yr, data = plot_data,
     pch = 16,
     xlab = "Year of release",
     ylab = expression("Yield ( Mg ha"^-1~")"),
     ylim = c(9,13),
     xlim = c(1984, 2018.5))

segments(x0 = plot_data[,"yr"],
         y0 = plot_data[,"mean"] + plot_data[,"se"],
         y1 = plot_data[,"mean"] - plot_data[,"se"])

#Used locator() it interactively find label points 
aa = readRDS("f2_text_pos.rds")
text(x = aa$x, y = aa$y,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.75)
#### Polygons
yrs <- 1975:2025

abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000,
       lty = 2)
polygon(x = c(yrs, rev(yrs)),
        y = c(ints2[1,],rev(ints2[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))

# text(x = plot_data$yr, y = plot_data$mean,
#      labels = 1:nrow(plot_data), #gsub("M", "M-", row.names(plot_data)),
#      pos = 4, cex = 0.75)
# mm = lm(yield_kg/1000 ~ release_yr, data = mod_data[idx,])
# abline(mm, lty = 2)
# yrs = 1980:2020
# int = predict(mm, newdata = data.frame(release_yr = yrs), se.fit = TRUE)

# polygon(x = c(yrs, rev(yrs)),
#         y = c(int$fit + int$se.fit,rev(int$fit - int$se.fit)),
#         border = NA,
#         col = rgb(0,0,0,0.1))

\caption{The yield of released medium grain rice varieties for only
the two years prior to each cultivar's official release. Points
($\bullet$) and bars (|) represent the mean and standard error of the
raw data, respectively. The dashed line and shaded area are the
genetic yield improvement and credible interval as estimated by the model.}
\label{fig:first_yrs}
\end{figure}

@@

Comparing the cultivars' yield during each's initial time in the
experimental trials confirmed that genetic yield improvements have
been consistant over the study period
(Fig. \ref{fig:first_yrs}). As estimated by the full model, there is
no conclusive evidence that the yields of medium grain rice cultivars
selected and released by CCRRF's breeding program have, on average,
increased over the last 30 years (Table \ref{tbl:coef}). The full
model which corrected for both the year of release and yield erosion
was estimated to have better out-of-sample predictive accuracy
compared to reduced models containing only one predictor. The
estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

When yield erosion was not corrected for in the comparison
(i.e., omitted from the model), there is an apparent small increase
per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). Although this
estimate ignores the confounding effect of yield erosion over time and
performed worse in out-of-sample predictive accuracy, it confirms that
yields in side-by-side comparison would support the conclusion that newer
cultivars out-preform older ones if yield erosion is not considered.

After correcting for the other variables, some cultivars were estimated
to have higher yield compared to the overall average (as estimated by
random intercepts). The marginal estimate for the performance of a
cultivar averaged over the uncertainty in the other parameters in the
model (i.e., the random effect) shows three cultivars with higher than
average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}). A fourth, M-209,
might also have higher than average yield, though the CI for M-209
overlaps with zero (Fig. \ref{fig:ids_comp}).

@@

@
\caption{The relationship between yield and the release year of medium
grain rice cultivars for the last 6 years
(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
correcting for yield erosion over time.}
\label{fig:wo_erosion}
\end{figure}

\begin{comment}
\begin{figure}[H]
<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=

@@

@
% Change from a./b. to accounting for erosion, More details about how
% yield was corrected - 
\caption{The relationship between yield and the release year of medium
grain rice cultivars for the last 6 years
(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
correcting for yield erosion over time. Yield erosion was corrected
for by adjusting each observed yield by estimated amount lost 
  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
\label{fig:w_erosion}
\end{figure}
\end{comment}

@@

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is estimated to be
  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
  and bars (|) represent the mean and standard error of the raw data,
  respectively. The shaded area is the 90\% credible interval.}
% For each year, point size by n (open circles)
\label{fig:RES_year}
\end{figure}

@
\caption{The performance of two major cultivars over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M-202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs}
\end{figure}

The average decrease in yield over time once a cultivar is selected
(i.e., yield erosion) was estimated as \Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to
\Sexpr{RES_CI[1]}). The oldest cultivar in the trials had lost an
average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since
its release. Comparing the two dominate medium grain rice cultivars M-202
(widely planted from the late 1980s until the early 2000s), and M-206
(the most widely planted CCRRFS cultivar at approximately 50\% of
planted area at the time of writing), these two cultivars experienced
similar trends in yield decline (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:coef}). The currently estimated
\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206 over M-202
can be attributed solely to the erosion of M-202's yield over time
(Figs \ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further
supporting this estimate of yield erosion, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:compare_yrs}, Table \ref{tbl:ids}).

@@

We do not find conclusive evidence of increased genetic yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}. Since the estimated yield improvement
spans from \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
year\tsups{-1}, both increases and decreases in the genetic yield
potential over the study period are plausible. However, even the
estimated yield improvement that does not account for yield erosion
(\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1};
Fig. \ref{fig:wo_erosion}) is much lower than previous estimated,
i.e., 37 to 60 kg ha\tsups{-1} year\tsups{-1} \citep{Omar2014}. Based on the
estimates from this study, there is less than a 1\% probability that
the true value is higher than 45 kg ha\tsups{-1}
year\tsups{-1}. Therefore, we conclude that the CCRRF breeding program
has likely been maintaining but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures.

We estimate that yield erosion has decreased the yield of medium grain
rice cultivars released by the CCRRF over the last 30+ years by an
average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially offsetting
any gains made by the crop improvement program. Biotic and abiotic forces have
previously been speculated to be responsible for yield erosion, since
the cultivar's genetics remain stable from the time of selection but
the conditions the cultivar may experience over time are continuously
changing \citep{Peng1998, Peng2000, Peng2010}. The trend observed here
is consistent with previous reports of the erosion of genetic gains of
cultivar IR8 at IRRI \citep{Peng1998, Peng2000}, suggesting the
erosion of yield over time is a phenomenon that widely affects rice
breeding programs. This observation highlights both the need for maintenance
of plant breeding and the difficulty of increasing genetic yield potential,
especially in systems such as CA where yield is approaching the
physiological limit \citep{espe16a}. Without maintenance a breeding
program and continuously selecting cultivars well adapted to current conditions,
CA rice production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
the study period.
